### 2024-03-23 thesis/input_files/probabilistic.tex
gradient boosting

% \section{Copula models}
\section{Quantile forest}
Meinshausen \cite{meinshausen2006quantile} extends the idea of random forest \cite{breiman2001random} generalising it, the result is the quantile forest algorithm. Quantile forest allows us to estimate conditional quantiles in a non parametric fashion.
\\
In order to understand this algorithm it is necessary to first cover the theory of decision trees and random forests.
\subsection{Decision trees}
Decision trees methods partition recursively the feature space in a set of binary rectangles and then fit a simple model in each of those partitions (the most straightforward is fitting just a constant).
To get started, we first split the space into two disjoint regions, then we model the response variable by the mean of the observed predicted variables with associated features falling in that specific region. Our goal is selecting the best features and best split point such that to achieve the best generalizing fit.
For an example/visualisation consider the pictures \ref{fig:elements_statistical_learning1} and \ref{fig:elements_statistical_learning2} taken from \cite{hastie2009elements}, where the decision tree algorithm is visualised for a regression problem with two independent variables X2 and X1.
\begin{figure}
    \includegraphics[width=0.5\textwidth]{images/elsii1.png}
    \caption{two-dimensional feature space partitioned by recursive binary splittin}
    \label{fig:elements_statistical_learning1}
  \end{figure}
\\
\begin{figure}
\includegraphics[width=1\textwidth]{images/elsii2.png}
\caption{partition tree and regression model}
\label{fig:elements_statistical_learning2}
\end{figure}
\\
Suppose to partition the feature space into M regions $R_1,\dots,R_M$, then the model reads as follows.
\begin{equation}
    f(x)=\sum\limits_{m=1}^{M}c_m\mathbb{1}(x \in R_m)
\end{equation}
\\
It follows that the function minimising the sum of squares is the one with $\hat{c}_m=mean(y_i|x_i \in R_m)$
Finding the best split in terms of minimum sum of squares is computationally infeasible in practice. Therefore we approximate a solution by approaching the problem in a greedy fashion.
Let $j$ denote the splitting variable, $s$ be the the split point we define the two half planes
\begin{equation}
    R_1(j,s)=\{X|X_j\leq s\} \quad R_2(j,s)=\{X|X_j\>s\}
\end{equation}
Then we search for the $s$ and $j$ that solve
\begin{equation}
    \min_{j,s}\left[\min_{c_1} \sum\limits_{x_i \in R_1(j,s)}(y_i-c_1)^2+\min_{c_2} \sum\limits_{x_i \in R_2(j,s)}(y_i-c_2)^2\right]
\end{equation}
The inner problem is easy, as already pointed out we will have 
\begin{equation}
    \hat{c}_1=mean(y_i|x_i \in R_1) \quad \hat{c}_2=mean(y_i|x_i \in R_2)
\end{equation}
For the outer problem, we scan through all the (j,s) tuples and pick the best pair. 
Next, one or both of these regions from the previous step are split into two more regions; we recurse this process until some stopping condition is triggered (max number of branches, max deth of tree, threshold on the MSE, minimum number of observation in each leaf node).
Notice, this being a greedy algorithm implies that our final solution is guaranteed to be just a local optimum not a global one.
Even though their simplicity, these models have proved themselves to be really powerful, see figure \ref{fig:decision_tree} for a visualisaiton . The most popular among these model is the CART \cite{breiman2017classification} tree, its name comes from the fact that it can handle both classification and regression problems.
\begin{figure}
    \includegraphics[width=\textwidth]{images/decision_tree.png}
    \caption{decision tree regression}
    \label{fig:decision_tree}
    % image generated on coolab to be quick
\end{figure}

@@

\subsection{Quantile forest}
The key observation here is noting that random forests approximates the conditional mean $\mathbb{E}(Y|X=x)$ by a weighted average over the observed $Y$. Hence, we can extend this idea to the full conditional distribution by
\begin{equation}
    F(y|X=x)=P(Y\leq y|X=x)=\mathbb{E}(\mathbb{1}_{\{Y\leq y\}}|X=x)
\end{equation}
All we have to do is approximating $\mathbb{E}(\mathbb{1}_{\{Y\leq y\}}|X=x)$ by a weighted mean over the random variable $\mathbb{1}_{Y\leq y}$
\begin{equation}
    \hat{F}(y|X=x)=\sum\limits_{i=1}^{n}\omega_i(x)\mathbb{1}_{\{Y_i\leq y\}}
\end{equation}

\section{Quantile gradient boosting machine}

@@

We end this section with the derivation of the dual problem; that is the convex problem we will solve in order to get the qunatiles prdeiction of our quantile kernel algorithm.

@@

Switching sign, we rephrase it as a minisation problem, which is the common practice in convex optimisation.

@@

In order to get acquainted with the inner working of the presented methods, this section covers an application explaining practical details and comparing results.

@@

- metrics pinball, picw plot, mae error

>>>>

% \section{Copula models}
\section{Quantile forest}
Meinshausen \cite{meinshausen2006quantile} extends the idea of random forest \cite{breiman2001random} generalising it, the result is the quantile forest algorithm. Quantile forest allows us to estimate conditional quantiles in a non parametric fashion.
\\
In order to understand this algorithm it is necessary to first cover the theory of decision trees and random forests.
\subsection{Decision trees}
Decision trees methods partition recursively the feature space in a set of binary rectangles and then fit a simple model in each of those partitions (the most straightforward is fitting just a constant).
To get started, we first split the space into two disjoint regions, then we model the response variable by the mean of the observed predicted variables with associated features falling in that specific region. Our goal is selecting the best features and best split point such that to achieve the best generalizing fit.
For an example/visualisation consider the pictures \ref{fig:elements_statistical_learning1} and \ref{fig:elements_statistical_learning2} taken from \cite{hastie2009elements}, where the decision tree algorithm is visualised for a regression problem with two independent variables X2 and X1.
\begin{figure}
    \includegraphics[width=0.5\textwidth]{images/elsii1.png}
    \caption{two-dimensional feature space partitioned by recursive binary splittin}
    \label{fig:elements_statistical_learning1}
  \end{figure}
\\
\begin{figure}
\includegraphics[width=1\textwidth]{images/elsii2.png}
\caption{partition tree and regression model}
\label{fig:elements_statistical_learning2}
\end{figure}
\\
Suppose to partition the feature space into M regions $R_1,\dots,R_M$, then the model reads as follows.
\begin{equation}
    f(x)=\sum\limits_{m=1}^{M}c_m\mathbb{1}(x \in R_m)
\end{equation}
\\
It follows that the function minimising the sum of squares is the one with $\hat{c}_m=mean(y_i|x_i \in R_m)$
Finding the best split in terms of minimum sum of squares is computationally infeasible in practice. Therefore we approximate a solution by approaching the problem in a greedy fashion.
Let $j$ denote the splitting variable, $s$ be the the split point we define the two half planes
\begin{equation}
    R_1(j,s)=\{X|X_j\leq s\} \quad R_2(j,s)=\{X|X_j\>s\}
\end{equation}
Then we search for the $s$ and $j$ that solve
\begin{equation}
    \min_{j,s}\left[\min_{c_1} \sum\limits_{x_i \in R_1(j,s)}(y_i-c_1)^2+\min_{c_2} \sum\limits_{x_i \in R_2(j,s)}(y_i-c_2)^2\right]
\end{equation}
The inner problem is easy, as already pointed out we will have 
\begin{equation}
    \hat{c}_1=mean(y_i|x_i \in R_1) \quad \hat{c}_2=mean(y_i|x_i \in R_2)
\end{equation}
For the outer problem, we scan through all the (j,s) tuples and pick the best pair. 
Next, one or both of these regions from the previous step are split into two more regions; we recurse this process until some stopping condition is triggered (max number of branches, max deth of tree, threshold on the MSE, minimum number of observation in each leaf node).
Notice, this being a greedy algorithm implies that our final solution is guaranteed to be just a local optimum not a global one.
Even though their simplicity, these models have proved themselves to be really powerful. See figure \ref{fig:decision_tree} for a example where decision tree with different hyperparameters are fitted to a sine wave plus noise. The most popular among these model is the CART \cite{breiman2017classification} tree, its name comes from the fact that it can handle both classification and regression problems.
\begin{figure}
    \includegraphics[width=\textwidth]{images/decision_tree.png}
    \caption{decision tree regression}
    \label{fig:decision_tree}
    % image generated on coolab to be quick
\end{figure}

@@

\subsection{Quantile forest}
The key observation here is noting that random forests approximates the conditional mean $\mathbb{E}(Y|X=x)$ by a weighted average over the observed $Y$. Hence, we can extend this idea to the full conditional distribution by
\begin{equation}
    F(y|X=x)=P(Y\leq y|X=x)=\mathbb{E}(\mathbb{1}_{\{Y\leq y\}}|X=x)
\end{equation}
All we have to do is approximating $\mathbb{E}(\mathbb{1}_{\{Y\leq y\}}|X=x)$ by a weighted mean over the random variable $\mathbb{1}_{\{Y\leq y\}}$
\begin{equation}
    \hat{F}(y|X=x)=\sum\limits_{i=1}^{n}\omega_i(x)\mathbb{1}_{\{Y_i\leq y\}}
\end{equation}
By swapping $F(y|X=x)$ with $\hat{F}(y|X=x)$ in the defintion of conditional quantiles we obtain their respective random forest estimator
\begin{equation}
    \hat{Q}_\alpha=\inf\{y:\hat{F}(y|X=x)\geq \alpha\}
\end{equation}

\section{Quantile gradient boosting machine}
\subsection{Boosting}
With boosting we fit an additive expansion of elementary basis functions; with M basis functions we have 
\begin{equation}
    f(x)=\sum\limits_{m=1}^{M}\beta_m b(x;\gamma_m)
\end{equation}
where $b(x;\gamma)$ are the basis functions while $\beta_m$ are the coefficients of the expansion.
Boosted models are fitted by minimizing a loss function over the training data
\begin{equation}
    \min_{\beta_m, \gamma_m}\sum\limits_{i=1}^{N}L\left(y_i, \sum\limits_{m=1}^M \beta_m b(x_i;\gamma_m)\right)
\end{equation}
However, such problem is highly intensive in terms of computation. Therefore, what is done in the literature is approximating its solution by iteratively adding new basis function to the current expansion. That is, we construct $f_m$ by solving for the optimal basis function and coefficients to add to $f_{m-1}$. Considering the square loss, we would have
\begin{equation}
    L(y_i, f_{m-1}(x_i)+\beta_m b(x_i;\gamma))=(e_{im}-\beta_m b(x_i;\gamma))^2
\end{equation}

\subsection{Boosted trees}
We combine several trees obtaining the boosted tree model
\begin{equation}
    f_M(x)=\sum\limits_{m=1}^M T(x;\Theta_m)
\end{equation}
Thus, at each step of the iterative optimisation procedure we have to solve
\begin{equation}\label{boosting_minimisation_problem}
    \hat{\Theta}_m=\argmin{\Theta_m}\sum\limits_{i=1}^{N}L(y_i, f_{m-1}(x_i)+T(x_i;\Theta_m))
\end{equation}
Remember, $\Theta_m$ referes to the parameters of the mth tree, $\Theta_m=\{R_{jm}, \gamma_{jm}\}_1^{J_m}$
\subsection{Gradient boosting}
In order to robustly solve \ref{boosting_minimisation_problem}, gradient boosting considers the following problem
\begin{equation}
    \tilde{\Theta_m}=\argmin{\Theta}\sum\limits_{i=1}^{N}\left(-g_{im}-T(x_i;\Theta)\right)^2
\end{equation}
where $g_{im}$ is the gradient of $L(f)=\sum\limits_{i=1}^N L(y_i, f(x_i))$ evaluated at $f_{m-1}$. Put simply, we are fitting the mth tree to the negative of the gradient values of f through least squares.
In order to solve our quantile regression tasks through gradient boosting, all we need to do is specifying the pinball loss as our criterion to guide the minimisation algorithm. See figure \ref{fig:gradient_boosting} for a visualisation
\begin{figure}
    \includegraphics[width=\textwidth]{images/gradient_boosting.png}
    \caption{gradient boosting regression}
    \label{fig:gradient_boosting}
    % image generated on coolab to be quick
\end{figure}

@@

We end this section with the derivation of the dual problem; that is the convex problem we will solve in order to get the quantiles prdeiction of our quantile kernel algorithm.

@@

Switching sign, we rephrase it as a minimisation problem, which is the common practice in convex optimisation.

@@

In order to get acquainted with the inner workings of the presented methods, this section covers an application explaining practical details and comparing results.

@@

- metrics pinball, picw plot, mae error on test data
